Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control: A Secondary Analysis of SPRINT

### 2022-02-08 doc/manuscript.Rmd
incorporate feedback from BL

Byron C. Jaeger, PhD,^1^ Adam P. Bress, PharmD, MS,^2,3^ Joshua D. Bundy, PhD,^4^ Alfred K. Cheung, MD,^5,6^ William C. Cushman, MD,^7^ Paul E. Drawz, MD,^8^ Lawrence J. Fine, MD,^9^ Karen C. Johnson, MD,^7^ Cora E. Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V. Rocco, MD,^12^ Stephen R. Rapp, PhD, ^13^ Joni K. Snyder, RN,BSN,^9^ Mark A. Supiano, MD,^14^ Paul K. Whelton, MD,^4^ Jeff D. Williamson, MD, MHS,^15^ Jackson T. Wright, Jr, MD, PhD,^16^ David M. Reboussin, PhD,^1^ and Nicholas M. Pajewski, PhD^1^

@@

@@

__OBJECTIVE__. To evaluate the legacy effect of receiving intensive treatment for approximately three years during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.

__DESIGN, SETTING, AND PARTICIPANTS__. Randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke. Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and trial follow-up visits occurred through July 2016.

@@

__MAIN OUTCOMES AND MEASURES__. Trial follow-up supplemented with mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020. In a subset of `r n_ehr` trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.

__RESULTS__. Among `r n_total` randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years. Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial. However, there was no evidence of benefit during post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`). The estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased post-trial, from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.

__CONCLUSIONS AND RELEVANCE__. The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during post-trial observational follow-up. Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.

@@

Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD) [@gdbJACC2020]. Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults and lowest CVD risk when systolic BP (SBP) is 120 to 124 mm Hg [@rahimi_age_stratified_2021; @bundy_systolic_2017]. The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality [@sprint_research_group_randomized_2015]. Because a clear benefit was evident for intensive treatment after a median follow-up of 3.3 years, SPRINT was stopped early. However, the persistence of benefit after the trial (i.e., the legacy effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].

The objective of the current study was to evaluate the legacy effect of intensive treatment during the SPRINT trial by analyzing post-trial all-cause and cardiovascular mortality. A secondary objective was to examine SBP following the conclusion of the trial follow-up. To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 to 2020 and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants.

@@

__Trial Design__: The design and methods of the SPRINT trial have been published previously [@ambrosius_design_2014; @sprint_research_group_randomized_2015]. Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD. Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed. Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older. Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease. Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site. Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent. The study is registered at ClinicalTrials.gov (NCT01206062).

@@

@@

tbl_object <- tbl_data |> 
  select(variable,
         group,
         .n_Standard_trial,
         .n_Intensive_trial,
         .i_Standard_trial,
         .i_Intensive_trial,
         .h_trial,
         .n_Standard_cohort,
         .n_Intensive_cohort,
         .i_Standard_cohort,
         .i_Intensive_cohort,
         .h_cohort) |> 
  as_grouped_data(groups = 'variable') |> 
  drop_na_rows() |> 
  as_flextable(hide_grouplabel = TRUE) |> 
  add_header_row(
    values = c(" ", 
               "N Events / N Total", 
               "Incidence\n(95% CI)", 
               "Hazard Ratio\n(95% CI)",
               "N Events / N Total", 
               "Incidence\n(95% CI)",
               "Hazard Ratio\n(95% CI)"),
    colwidths = c(1, 2, 2, 1, 2, 2, 1)
  ) |> 
  add_header_row(
    values = c(" ", 
               "Trial Phase",
               "Cohort Phase"),
    colwidths = c(1, 5, 5)
  ) |> 
  set_header_labels(
    group = " ",
    .n_Standard_trial   = "Standard",
    .n_Intensive_trial  = "Intensive",
    .i_Standard_trial   = "Standard",
    .i_Intensive_trial  = "Intensive",
    .h_trial            = "Hazard Ratio\n(95% CI)",
    .n_Standard_cohort  = "Standard",
    .n_Intensive_cohort = "Intensive",
    .i_Standard_cohort  = "Standard",
    .i_Intensive_cohort = "Intensive",
    .h_cohort           = "Hazard Ratio\n(95% CI)"
  ) |> 
  merge_v(part = 'header') |>
  theme_box() |> 
  padding(j = 1, i = ~ !is.na(group), padding.left = 15, part = 'body') |> 
  bg(i = ~ is.na(group), bg = 'grey80') |> 
  italic(i = ~ is.na(group), italic = TRUE) |> 
  width(width = 0.95) |> 
  width(j=1, width = 1.5) |> 
  align(align = 'center', part = 'all') |> 
  align(j = 1, align = 'left', part = 'all') |> 
  fontsize(size = 12, part = 'all') |> 
  font(fontname = 'Times New Roman', part = 'all') |> 
  footnote(i = ~ variable == 'Chronic Kidney Disease', 
           j = 1,
           part = 'body',
           ref_symbols = 'a',
           value = ckd_defn) |> 
  footnote(i = 1, 
           j = 1, 
           ref_symbols = '',
           value = abbrvs_write(abbrvs[c('CI', 'FI')]))

>>>>

Byron C. Jaeger, PhD,^1^ Adam P. Bress, PharmD, MS,^2,3^ Joshua D. Bundy, PhD,^4^ Alfred K. Cheung, MD,^5,6^ William C. Cushman, MD,^7^ Paul E. Drawz, MD,^8^ Lawrence J. Fine, MD,^9^ Karen C. Johnson, MD,^7^ Cora E. Lewis, MD, MSPH^10^ Suzanne Oparil, MD,^11^ Michael V. Rocco, MD,^12^ Stephen R. Rapp, PhD, ^13^ Joni K. Snyder, RN,BSN,^9^ Mark A. Supiano, MD,^14^ Paul K. Whelton, MD,^4^ Jeff D. Williamson, MD, MHS,^15^ Jackson T. Wright, Jr, MD, PhD,^16^ David M. Reboussin, PhD,^1^ and Nicholas M. Pajewski, PhD^1^

@@

<!-- TODO: INKSCAPE A TIMELINE WITH TRIAL PHASES -->

@@

__OBJECTIVE__. To evaluate the legacy effect of receiving intensive treatment during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.

__DESIGN, SETTING, AND PARTICIPANTS__. Randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke. Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and observational trial follow-up visits occurred through July 2016.

@@

__MAIN OUTCOMES AND MEASURES__. Extended trial follow-up assessed mortality via the US National Death Index, beginning in 2016 through December 31st, 2020. In a subset of `r n_ehr` trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.

__RESULTS__. Among `r n_total` randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years. Intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through trial close-out visits. However, there was no evidence of benefit during extended post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`). The estimated mean (95% CI) outpatient SBP among participants randomized to intensive treatment increased from `r f(bp_viz_5yr$Intensive)` at five years to `r f(bp_viz_10yr$Intensive)` ten years post-randomization.

__CONCLUSIONS AND RELEVANCE__. The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during extended post-trial follow-up. Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.

@@

Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD) [@gdbJACC2020]. The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a SBP target < 120 mm Hg, reduced the risk for cardiovascular and all-cause mortality [@sprint_research_group_randomized_2015]. Because a clear benefit was evident for intensive treatment after a median follow-up of 3.3 years, SPRINT was stopped early. However, the long-term persistence of benefit after the trial (i.e., the legacy effect) of intensive treatment on cardiovascular and all-cause mortality has not been evaluated [@wander_legacy_2018].

The objective of the current study was to evaluate the long-term legacy effect of intensive treatment during the SPRINT trial by analyzing post-trial all-cause and cardiovascular mortality. A secondary objective was to examine SBP following the conclusion of the trial follow-up. To accomplish these objectives, we linked participants to the National Death Index (NDI) from 2016 through 2020 and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) for a subset of trial participants.

@@

__Trial Design__: The design and methods of the SPRINT trial have been published previously [@ambrosius_design_2014; @sprint_research_group_randomized_2015]. Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD. Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed. Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^) [@levey_new_2009], or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older. Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease. Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site. Randomization began on November 8, 2010 and ended in March 2013. The trial intervention ended on August 20, 2015, and observational trial follow-up visits occurred through July 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent. The study is registered at ClinicalTrials.gov (NCT01206062).

@@

Meta analyses of randomized trials have shown that pharmacological blood pressure (BP) lowering reduces the risk of major CVD events across the spectrum of initial BP [@rahimi_age_stratified_2021; @bundy_systolic_2017].

@@

tbl_object <- tbl_data |> 
  select(variable,
         group,
         .n_Standard_trial,
         .n_Intensive_trial,
         .i_Standard_trial,
         .i_Intensive_trial,
         .h_trial,
         .n_Standard_cohort,
         .n_Intensive_cohort,
         .i_Standard_cohort,
         .i_Intensive_cohort,
         .h_cohort) |> 
  as_grouped_data(groups = 'variable') |> 
  drop_na_rows() |> 
  as_flextable(hide_grouplabel = TRUE) |> 
  add_header_row(
    values = c(" ", 
               "N Events / N Total", 
               "Incidence\n(95% CI)", 
               "Hazard Ratio\n(95% CI)",
               "N Events / N Total", 
               "Incidence\n(95% CI)",
               "Hazard Ratio\n(95% CI)"),
    colwidths = c(1, 2, 2, 1, 2, 2, 1)
  ) |> 
  add_header_row(
    values = c(" ", 
               "Trial Phase",
               "Cohort Phase"),
    colwidths = c(1, 5, 5)
  ) |> 
  set_header_labels(
    group = " ",
    .n_Standard_trial   = "Standard",
    .n_Intensive_trial  = "Intensive",
    .i_Standard_trial   = "Standard",
    .i_Intensive_trial  = "Intensive",
    .h_trial            = "Hazard Ratio\n(95% CI)",
    .n_Standard_cohort  = "Standard",
    .n_Intensive_cohort = "Intensive",
    .i_Standard_cohort  = "Standard",
    .i_Intensive_cohort = "Intensive",
    .h_cohort           = "Hazard Ratio\n(95% CI)"
  ) |> 
  merge_v(part = 'header') |>
  theme_box() |> 
  padding(j = 1, i = ~ !is.na(group), padding.left = 15, part = 'body') |> 
  bg(i = ~ is.na(group), bg = 'grey80') |> 
  italic(i = ~ is.na(group), italic = TRUE) |> 
  width(width = 0.95) |> 
  width(j=1, width = 1.5) |> 
  align(align = 'center', part = 'all') |> 
  align(j = 1, align = 'left', part = 'all') |> 
  fontsize(size = 12, part = 'all') |> 
  font(fontname = 'Times New Roman', part = 'all') |>
  footnote(i = 2, j = 4,
           part = 'header',
           ref_symbols = 'a',
           value = as_paragraph('Incidence is presented as the expected rate of events per 1,000 person-years.')) |> 
  footnote(i = ~ variable == 'Chronic Kidney Disease', 
           j = 1,
           part = 'body',
           ref_symbols = 'b',
           value = ckd_defn) |> 
  footnote(i = 1, 
           j = 1, 
           ref_symbols = '',
           value = abbrvs_write(abbrvs[c('CI', 'FI')]))
